Anomaly detection with kernel density estimation on manifolds

### 2022-07-05 paper/paper.tex
Rewrite kde for riem manifolds

\makeatletter
\@ifpackageloaded{biblatex}{}{\usepackage[style=authoryear-comp, backend=biber, natbib=true]{biblatex}}
\makeatother
\ExecuteBibliographyOptions{bibencoding=utf8,minnames=1,maxnames=3, maxbibnames=99,dashed=false,terseinits=true,giveninits=true,uniquename=false,uniquelist=false,doi=false, isbn=false,url=true,sortcites=false, date=year}

@@

%% Any special functions or other packages can be loaded here.
\usepackage{tabu,placeins,algorithmic}
\usepackage[ruled,vlined,linesnumbered]{algorithm2e}
\urlstyle{tt}  % use monospace font for urls
\usepackage[cal = txupr]{mathalpha}
\geometry{margin=2.2cm}
\mathtoolsset{showonlyrefs}
\def\sectionautorefname{Section}
\def\subsectionautorefname{Section}
\usepackage{bm}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

% Adjust headwidth in case user has changed geometry in header-includes
\renewcommand{\headwidth}{\textwidth}

@@

\hypertarget{vkde}{%
\section{Kernel density estimation on manifolds}\label{vkde}}

@@

\hypertarget{multivariate-kernel-density-estimation}{%
\subsection{Multivariate kernel density estimation}\label{multivariate-kernel-density-estimation}}

@@

For computational purposes, \textcite{Duong2007-up} implements the kernel density estimates for data from 1 to 6 dimensions in the R package \emph{ks}.
However, for variable kernel density estimation where the bandwidth is no longer a constant matrix, the \emph{ks} implementation is limited to 2-dimension and the bandwidth can either be a diagonal matrix or a vector of the diagonal elements, which limits the full flexibility of the kernel function.
More importantly, the estimator in Equation \eqref{eq:vkde} is only applicable when the sample space is Euclidean. In the cases of a Riemannian manifold, the underlying density is supported on a single manifold. The fixed bandwidth matrix assumes that the smoothing is the same across the entire manifold and there is no local distortion at each sample point. This motivates our proposed adaptive density estimator for Riemannian manifolds as described in \autoref{riemest}.

@@

Consider a compact Riemannian manifold \((M, g)\) of dimension \(d\) without boundary and a probability distribution with density \(f\) on the manifold. Assume \((M, d_g)\) is a complete metric space, where \(d_g\) is the Riemannian distance induced by the Riemannian, \(g\), and the strictly positive injectivity radius \autocite{Chavel2006-mp} of the manifold, \(inj_g(M)\). Denote \(\pmb{X}_i, i = 1, 2, \dots, N\), where \(\pmb{X}_i \in \mathcal{R}^d\), as i.i.d. random objects on \(M\) with density \(f\). For each point \(p\in M\), \textcite{Pelletier2005-vu} proposed a kernel density estimator of \(f\) to be
\begin{equation}
\label{eq:denriem}
\hat{f}(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{h^d \theta_{\pmb{X}_i}(p)} K(\frac{d_g(p, \pmb{X}_i)}{h}),
\end{equation}
where \(h\) is the global bandwidth satisfying \(h \leq h_0\) for fixed \(h_0\) such that \(0\leq h_0\leq inj_g(M)\), \(K: \mathcal{R}_{+} \rightarrow \mathcal{R}\) is a non-negative map, and \(\theta_{q}(p)\) is the volume density function on the manifold.
\textcite{Pelletier2005-vu} mentions that the volume density function \(\theta_q(p)\) defined for \(p\) in a neighborhood of \(q\) on the manifold, with the geodesic normal coordinates at \(q\), is equal to the determinant of the Riemannian metric \(g\) at the logarithm map at q, \(\exp_q^{-1}(p)\). The expression in \eqref{eq:denriem} is also proved to be consistent with the kernel density estimators in the Euclidean case as
\begin{equation}
\label{eq:denconsist}
\hat{f}(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{h^d} K(\frac{\|p-\pmb{X}_i\|}{h}),
\end{equation}
and converges at the same rate as the Euclidean kernel density estimator \autocite{Henry2009-ll}.

\hypertarget{volume-density-function}{%
\subsubsection{Volume density function}\label{volume-density-function}}

@@

\hypertarget{volume-density-function-1}{%
\subsection{Volume density function}\label{volume-density-function-1}}

>>>>

\makeatletter
\@ifpackageloaded{biblatex}{}{\usepackage[style=authoryear-comp, backend=biber, natbib=true]{biblatex}}
\makeatother
\ExecuteBibliographyOptions{bibencoding=utf8,minnames=1,maxnames=3, maxbibnames=99,dashed=false,terseinits=true,giveninits=true,uniquename=false,uniquelist=false,doi=false, isbn=false,url=true,sortcites=false}

@@

%% Any special functions or other packages can be loaded here.
\usepackage{tabu,placeins,algorithmic}
\usepackage[ruled,vlined,linesnumbered]{algorithm2e}
\urlstyle{tt}  % use monospace font for urls
\usepackage[cal = txupr]{mathalpha}
\geometry{margin=2.2cm}
\mathtoolsset{showonlyrefs}
\def\sectionautorefname{Section}
\def\subsectionautorefname{Section}
\usepackage{bm}
\DeclareMathOperator\supp{supp}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

@@

\hypertarget{vkde}{%
\section{Kernel density estimation on manifold learning embeddings}\label{vkde}}

@@

\hypertarget{mvkde}{%
\subsection{Multivariate kernel density estimation}\label{mvkde}}

@@

For computational purposes, \textcite{Duong2007-up} implements the kernel density estimates for data from 1 to 6 dimensions in the R package \emph{ks}.
However, for variable kernel density estimation where the bandwidth is no longer a constant matrix, the \emph{ks} implementation is limited to 2-dimension and the bandwidth can either be a diagonal matrix or a vector of the diagonal elements, which limits the full flexibility of the kernel function.
More importantly, the estimator in Equation \eqref{eq:vkde} is only applicable when the sample space is Euclidean. In the cases of a Riemannian manifold, the manifold has a differentiable structure, which has to be taken into account when considering the neighborhood of a sample point. Therefore, we need to consider the probability density estimation on a Riemannian manifold and adapts the techiniques on Euclidean samples to a non-Euclidean setting. This is discussed in \autoref{riemest}.

@@

Consider a compact Riemannian manifold \((M, g)\) of dimension \(d\) without boundary and a probability distribution with density \(f\) on the manifold. Assume \((M, d_g)\) is a complete metric space, where \(d_g\) is the geodesic distance induced by the Riemannian metric \(g\) and the strictly positive injectivity radius of the manifold \(inj_g(M)\) \autocite{Chavel2006-mp}. Denote \(\pmb{X}_i, i = 1, 2, \dots, N\), where \(\pmb{X}_i \in \mathcal{R}^d\), as i.i.d. random objects on \(M\) with density \(f\). To estimate the density of the manifold, \textcite{Pelletier2005-vu} proposed a kernel density estimator for \(f\) at point \(\pmb{p}\in M\) as
\begin{equation}
\label{eq:denriem}
\hat{f}(\pmb{p}) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r^d \theta_{\pmb{X}_i}(\pmb{p})} K(\frac{d_g(\pmb{p}, \pmb{X}_i)}{r}),
\end{equation}
where \(r\) is the global bandwidth satisfying \(r \leq r_0\) for fixed \(r_0\) such that \(0\leq r_0\leq inj_g(M)\), \(K: \mathcal{R}_{+} \rightarrow \mathcal{R}\) is a non-negative map, and \(\theta_{\pmb{X}_i}(\pmb{p})\) is the volume density function on the manifold for point \(\pmb{X}_i\). The estimator in \eqref{eq:denriem} is also proved to be consistent with the kernel density estimators in the Euclidean case in Equation \eqref{eq:kde} and converges at the same rate as the Euclidean estimator \autocite{Henry2009-ll}.

Denote \(\lambda\) as the Lebesque measure of \(\mathbb{R}^d\). The kernels \(K\) in \eqref{eq:denriem} are functions of the geodesic distance of the manifold \(d_g\) and they satisfy the following conditions \autocite{Pelletier2005-vu}:
\begin{align}
& (i) \int_{\mathbb{R}^{d}} K(\|x\|) \mathrm{d} \lambda(x)=1;
(ii) \int_{\mathbb{R}^{d}} x K(\|x\|) \mathrm{d} \lambda(x)=0;
(iii) \int_{\mathbb{R}^{d}}\|x\|^{2} K(\|x\|) \mathrm{d} \lambda(x)<\infty; \\
& (iv) \supp K=[0; 1];
(v) \sup K(x)=K(0).
\end{align}
Note that conditon (iv) implies that the support of the kernel is between 0 and 1, which is guaranteed by the condition on the bandwidth, \(r\leq inj_g(M)\). Under such condition, for each \(\pmb{p} \in M\), there exists a normal coordinate neighborhood of \(\pmb{p}\) containing an open ball \(B(\pmb{p}, r) \in T_pM\) centered at \(\pmb{p}\) with radius \(r\).

Now consider the volume density function \(\theta_{\pmb{X}_i}(\pmb{p})\) in the estimator.
For any tangent vector \(u \in T_pM\), let \(\gamma(u): t \rightarrow \exp_p(\frac{tu}{\|u\|})\) denotes a curve passing through \(u\), and let \(w_i, i=1,\dots,d\) be Jacobi fields along \(\gamma\) such that \(w_i(0)=0\) for all \(i=1,\dots,d\) and \(\frac{Dw_1}{dt}(0)=\frac{u}{\|u\|}\). Then \(\frac{Dw_1}{dt}(0)\) forms an orthonomal basis of the tangent space \(T_pM\).
The volume density function on \(M\), \(\theta_p(q): T_pM \rightarrow \mathbb{R}\) is defined as
\[
\theta_p(u) = \|u\|^{d-1} \det(w_1(\|u\|),\dots,w_d(\|u\|)).
\]

For the exponential map \(\exp_p: T_pM \rightarrow M\) induced by pullback of a volume form \(\exp_p^*\textit{Vol}\) on \(T_pM\), \(\theta_p\) is its density with respect to the Lebesgue measure of the Euclidean structure on \(T_pM\), which gives
\[
d\exp_p^*\textit{Vo}l(u) = \theta_p(u)du,
\]
in normal coordinates. The volume density function is motivated by how a measure \(du\) changes under the change of coordinates. It is related to the integration on manifolds and plays an important part in studing the density of manifolds.
\textcite{Pelletier2005-vu} mentions that the volume density function \(\theta_q(p)\) defined for \(p\) in a neighborhood of \(q\) on the manifold, with the geodesic normal coordinates at \(q\), is equal to the determinant of the Riemannian metric \(g\) expressed in the coordinates at the logarithm map of q, \(\exp_q^{-1}(p)\).

\hypertarget{dckde}{%
\section{Distotion-corrected kernel density estimator (DC-KDE)}\label{dckde}}

@@

\hypertarget{volume-density-function-move-to-section-2.2.1}{%
\subsection{Volume density function {[}MOVE TO SECTION 2.2.1{]}}\label{volume-density-function-move-to-section-2.2.1}}
